---
title: "Cleland Introduction to R"
author: "Chuck Cleland"
date: "2/1/2020"
output:
  html_document:
    highlight: tango
    theme: spacelab
    toc: yes
    toc_depth: 5
  word_document:
    toc: yes
    toc_depth: '5'
always_allow_html: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning=FALSE, comment=NA, fig.width = 6, out.width = "90%", fig.align = "center", dev='svg')
```

# About me
* Senior Research Scientist and Biostatistician
* Center for Drug Use and HIV Research
* Research and Methods Interests
     + HIV prevention and care 
     + health disparities
     + meta-analysis
     + longitudinal data analysis
* Contact
     + email: cmc13@nyu.edu
     + phone: 212-992-9417
     + location: 433 First Ave., Room 737
   
# Goals for Today
* What is R
* Why I use R and you should too
* Getting data into and out of R
* Data management with R
* Making plots with R
* Common statistical analyses done in R (e.g., t-tests, regression)
* R packages
* Resources to learn more and the possibility of an ongoing R interest group

# What is R?
* A statistical computing language and environment
* Windows, Mac, Unix, Linux
* Open source
* Organized into packages of functions and other objects

# R vs. R Studio

# Basic Operations
```{r }
1 + 2 + 3
1 + 2 * 3
(1 + 2) * 3
c(0,1,2,3,5,8)
1:11
1:11 + 3
```

# Assignment
```{r}
x <- 3
y <- 1:10
z <- c(1,20,3,6)
```

# Evaluation
```{r}
x
y
z
x + y
x * z
```

# Getting help within R
```{r echo=TRUE, eval=FALSE}
?mean
help(mean)
```

# Types of R Objects
* Data frame
     + Like the typical two-dimensional table of data in other programs
* Vector
* matrix
     + Also a two dimensional table of data, but every variable is of the same type
* List
     + A way to combine objects, possibly of different types, into a composite object
     + Very powerful because you can apply functions to lists
     + Provides ways to get a great deal accomplished with very concise code
* Function

# Types of R variables in Data Frames
* Character
```{r }
my.character <- c('Emma','Olivia','Sophia')
my.character
```
* Date
```{r }
my.dates <- c(as.Date("2012-04-05"), Sys.Date())
my.dates
```
* Numeric
```{r }
my.numeric <- c(12,62.3,7,101)
my.numeric
```
* Factor
```{r }
my.factor <- factor(c(0,0,0,1,1), labels=c('Female','Male'))
my.factor
```
* Logical
```{r }
my.logical <- c(TRUE,FALSE,FALSE)
my.logical
```

# Vectorized Arithmetic
```{r }
weight.kg <- c(60, 72, 57, 90, 95, 72)
height.meters <- c(1.75, 1.80, 1.65, 1.90, 1.74, 1.91)
bmi <- weight.kg/height.meters^2
bmi
```
* Works element by element in the weight.kg and height.meters vectors
* There is no need for a loop

# Make simple R objects
```{r }
byear <- c(68,69,71,72)
byear
Boys <- data.frame(kid = c('JHC','CMC','REC','WBC'), 
                   byear = c(68,69,70,71))
Boys
```

# Access names and types of variables in a data frame
```{r }
names(Boys)
str(Boys)
```

# Accessing variables in data frames
```{r }
Boys$kid
with(Boys, mean(byear))
```

# The R Workspace
* Objects you create are stored in a workspace
* R allows working with multiple objects at once, not just a single dataset
* Here is how to show what is in the workspace at any given point:

```{r }
ls()
```

* When you load packages, the objects in those packages are in different workspaces
* I am going to load a package called haven which has functions for reading data from and writing data to other formats

```{r }
library(haven)
```
* I just made the objects in the haven package available
* I can see what those objects are (all functions) using a variation of the ls() command:
```{r }
ls("package:haven")
```

# Getting data into R
* read.table for plain text (comma or tab delimited)
* read_sav for SPSS (haven package)
* read_dta for Stata (haven package)
* read.xlsx for MS-Excel (openxlsx package)
* Many other formats and connections to databases

# read.table
```{r }
mytable <- read.table("c:/chuck/NYU/Stat Group/table1.txt", sep=",", header=TRUE)
mytable
str(mytable)
```

# read_sav
```{r }
library(haven)
titanic <- read_sav("c:/chuck/NYU/Stat Group/titanic.sav")
head(titanic)
attributes(titanic)$variable.labels
```

# Read and combine many files
* There are 8 MS-Excel files with the same structure in a folder
* Each has 10 observations on two variables, X and Y
* First, I load the package needed for reading Excel files, then get a list of the file names
```{r }
library(openxlsx)
my.files <- list.files("c:/chuck/NYU/Stat Group/R Workshop/Many Files/", 
                       pattern = ".xlsx$", full.names=TRUE, recursive=TRUE)
my.files # vector of the file names with full path
```

* Next I gather the names of the worksheets within each workbook
```{r }
my.sheets <- lapply(my.files, function(x){getSheetNames(x)}) # list of worksheet names 
                                                             # within each MS-Excel workbook
unlist(my.sheets)
```
* I put the workbook filenames and worksheet names together in a data frame
```{r }
my.sheets.DF <- data.frame(file = rep(my.files, unlist(lapply(my.sheets, length))), sheet = unlist(my.sheets))
my.sheets.DF
```
* I create an empty list to hold each data file
```{r }
my.list <- vector(mode="list", length = nrow(my.sheets.DF)) # empty list to hold each data file
```
* I loop over the file and worksheet names, reading each in and putting it into the list
```{r }
for(i in 1:nrow(my.sheets.DF)){
  my.list[[i]] <- read.xlsx(xlsxFile = as.character(my.sheets.DF$file)[i], 
                            sheet = as.character(my.sheets.DF$sheet)[i])  
}
```
* Finally, I combine the 8 data frames into one big data frame
```{r }
my.DF <- as.data.frame(do.call(rbind, my.list))
dim(my.DF) # 80 observations, 2 variables 
```
* The example has just 8 files in the folder, but it could be thousands and the principle is the same

# Getting data out of R
* write.table for plain text (comma or tab delimited)
* write_sav for SPSS (haven package)
* write_dta for Stata (haven package)
* write.xlsx for MS-Excel (openxlsx package)

# Data management: Recoding
```{r results='hide'}
library(car)
auto <- read_sav("c:/chuck/NYU/Stat Group/R Workshop/auto.sav")
summary(auto$MPG)
auto$MPG3 <- car::recode(auto$MPG, "lo:18=1; 19:23=2; 24:hi=3")
```

```{r }
table(auto$MPG3)
```

# Data management: Merging
Two data frames with a common ID variable
```{r }
A <- data.frame(famid = c(2,1,3), 
                name = c('Art','Bill','Paul'), 
                inc = c(22,30,25))
B <- data.frame(famid = c(3,1,2), 
                faminc96 = c(75,40,45), 
                faminc97 = c(76,40.5,40.4), 
                faminc98 = c(77,41,45.8))
intersect(names(A), names(B))
```

One Merged Data Frame
```{r }
A
B
AB <- merge(A, B)
AB
```

# Data management: Subsetting
```{r }
bank <- read_sav("c:/chuck/NYU/Stat Group/bank.sav")
head(bank)
```
* The <code>head()</code> function is a convenient way to see the first few observations of a data object

* With the <code>subset()</code> function

```{r }
subset(bank, SALBEG < 4000 & SEX == "Males", select=c(ID,SEX,SALBEG,AGE,SALNOW))
```

* Or by indexing
```{r }
bank[bank$SALBEG < 4000 & bank$SEX == "Males", c(1,3,2,5,6)]
```
* In both cases, I selected males with a beginning salary less than 4000, and only certain variables
     + subset specified the variables to include by name
     + indexing specified the variables to include based on their location in the data frame

# Data management: Reshaping, Wide to Long
```{r }
mytable <- read.table("c:/chuck/NYU/Stat Group/table1.txt", sep=",", header=TRUE)
mytable
library(dplyr)
library(tidyr)
longtab <- gather(mytable, year, inc, inc80:inc82)
longtab
```

# Data management: Reshaping, Long to Wide
```{r }
longtab
widetab <- spread(longtab, year, inc)
widetab
```

# Reshaping with data.table
* https://cran.r-project.org/web/packages/data.table/vignettes/datatable-reshape.html
* https://github.com/Rdatatable/data.table/wiki/Getting-started

# Data management: Aggregating
```{r }
head(iris)
aggregate(iris[,1:4], list(Species = iris$Species), FUN = mean)
```

# Data Management: Set Operations
```{r }
x <- c('A','B','C','D')
y <- c('D','A','F','J')
setdiff(x, y) # which elements of x are not in y
setdiff(y, x) # which elements of y are not in x
intersect(x, y) # what elements are in both x and y
```

```{r }
union(x, y) # elements in either x or y
"C" %in% x # is there a "C" in x
y %in% c("F","J") # which elements of y are equal to "F" or "J"
```

# Data Management: Character Manipulation
```{r }
x <- c('A','B','C','D')
grep('C', x)
gsub('C', 'c', x)
paste('Item', 1:10, sep="-")
x <- c('fruit_apple','instrument_guitar','instrument_piano','fruit_orange','instrument_trumpet')
grepl("fruit", x)
```

# Data Management: dplyr package
* <code>library(dplyr)</code>
* <code>filter()</code>
* <code>select()</code>
* <code>mutate()</code>
* <code>summarize()</code>
* <code>group_by()</code>
* <code>arrange()</code>

# Houston flights data
* All flights departing from Houston airports IAH and HOU
* Research and Innovation Technology Administration at the Bureau of Transporation Statistics

```{r }
library(hflights)
head(hflights)
```

```{r }
str(hflights)
```

# Using filter() to select observations
* January flights by American Airlines
```{r }
head(filter(hflights, Month == 1, UniqueCarrier == "AA"))
```

# Using select() to select variables

* Just the Departure Time, Arrival Time, and Air Time variables
```{r }
head(select(hflights, DepTime, ArrTime, AirTime))
```
* All variables that end with Time
```{r }
head(select(hflights, ends_with("Time")))
```

# Using mutate() to create new variables

```{r }
g1 <- mutate(hflights, ActualGroundTime = ActualElapsedTime - AirTime)
head(select(g1, ActualElapsedTime, AirTime, ActualGroundTime))
```

# Using summarize()

```{r }
summarize(hflights, min_dist = min(Distance),
                    mean_dist = mean(Distance),
                    median_dist = median(Distance),
                    sd_dist = sd(Distance),
                    max_dist = max(Distance))
```

# Using group_by() and summarize() to summarize by group
* This also illustrates the use of pipes in R
```{r }
hflights %>% group_by(UniqueCarrier) %>%
summarize(min_dist = min(Distance),
          mean_dist = mean(Distance),
          median_dist = median(Distance),
          sd_dist = sd(Distance),
          max_dist = max(Distance))
```

# Notice group_by() and summarize() essentially aggregate

```{r }
iris %>% 
  group_by(Species) %>%
  summarize(Sepal.Length = mean(Sepal.Length),
            Sepal.Width = mean(Sepal.Width),
            Petal.Length = mean(Petal.Length),
            Petal.Width = mean(Petal.Width))
```
* But much more flexibly!

# Putting data into a desired order with arrange()

```{r }
hflights %>% group_by(UniqueCarrier) %>%
summarize(min_dist = min(Distance),
          mean_dist = mean(Distance),
          median_dist = median(Distance),
          sd_dist = sd(Distance),
          max_dist = max(Distance)) %>%
arrange(mean_dist)
```

# Benefits of Piping

Without Piping
```{r}
arrange(
  summarize(
    group_by(
      filter(mtcars, carb > 1),
    cyl),
  Avg_mpg = mean(mpg)
  ),
desc(Avg_mpg)  
  )
```


With Piping
```{r}
mtcars %>%
  filter(carb > 1) %>%
  group_by(cyl) %>%
  summarize(Avg_mpg = mean(mpg)) %>%
  arrange(desc(Avg_mpg))
```

# Plots: Histogram
```{r fig.align='center', fig.width=8, fig.height=5}
hsb2 <- read_sav("https://clelandcm.github.io/Hands-On-R-Practice/hsb2.sav")
library(ggplot2)
ggplot(hsb2, aes(x = WRITE)) + geom_histogram(color = "grey", fill = "#34ebbd")
```

# Plots: Bar
```{r }
library(openxlsx)
fname <- read.xlsx("c:/chuck/NYU/Stat Group/R Workshop/Female-Names.xlsx", sheet = 1)
fname
```

```{r fig.align='center', fig.width=6, fig.height=6}
ggplot(fname, aes(x = Name, y = Ngirls)) + 
  geom_bar(stat = 'identity') + 
  ylab('Number') + xlab('') + 
  ggtitle('Ten Most Common Female Baby\nNames in 2014') +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))
```

# Plots: Scatter
ggplot2 package
```{r fig.align='center', fig.width=6, fig.height=6}
ggplot(hsb2, aes(x = WRITE, y = READ)) + geom_point()
```

Base R
```{r fig.align='center', fig.width=6, fig.height=6}
plot(READ ~ WRITE, data = hsb2)
```

# Plots: Scatter Enhanced
```{r fig.align='center', fig.width=8, fig.height=5}
library(ggplot2)
bank$GENDER <- as_factor(bank$GENDER)
ggplot(subset(bank, SALNOW < 30000), aes(x = SALBEG, y = SALNOW, color = GENDER)) + 
  geom_point()
```

# Plots: Scatter with Facets
```{r fig.align='center', fig.width=8, fig.height=5}
library(ggplot2)
ggplot(subset(bank, SALNOW < 30000), aes(x = SALBEG, y = SALNOW)) + 
  geom_point() + 
  facet_wrap( ~ GENDER)
```

# Plots: Boxplots
* Writing score on vertical axis
```{r fig.align='center', fig.width=3, fig.height=6}
hsb <- read_dta('c:/chuck/NYU/Stat Group/R Workshop/hsb.dta')
hsb$female <- as.factor(hsb$female)
ggplot(hsb, aes(x = female, y = write)) + geom_boxplot() + xlab("")
```

* Flip coordinates so writing score is on the horizontal axis
```{r fig.align='center', fig.width=8, fig.height=3}
ggplot(hsb, aes(x = female, y = write)) + geom_boxplot() + xlab("") + coord_flip()
```

# Plots: Alpha Transparency
```{r echo=FALSE, results='hide'}
myDF <- data.frame(X = runif(6000))
myDF$Y <- 100 + 2.43*myDF$X + rnorm(6000)
```

* Without transparency
```{r fig.align='center', fig.width=6, fig.height=6}
ggplot(myDF, aes(x=X,y=Y)) + geom_point(color="blue", size=3)
```

* With transparency
```{r fig.align='center', fig.width=6, fig.height=6}
ggplot(myDF, aes(x=X,y=Y)) + geom_point(color="blue", size=3, alpha =.20)
```

# Plots: Networks
```{r fig.align='center', fig.width=8, fig.height=5}
library(igraph)
net.bg <- barabasi.game(80) 
V(net.bg)$frame.color <- sample(c("white","black"), size = 80, replace=TRUE, prob=c(0.5,0.5))
V(net.bg)$color <- sample(c("orange","lightblue"), size = 80, replace=TRUE, prob=c(0.5,0.5))
V(net.bg)$label <- "" 
V(net.bg)$size <- 10
E(net.bg)$arrow.mode <- 0
plot(net.bg, layout=layout.lgl)
```

# Plots: More customized
```{r fig.width = 9, fig.height = 6, warning=FALSE, cache=TRUE}
library(ggplot2) # devtools::install_github("hadley/ggplot2")
library(ggalt)   # devtools::install_github("hrbrmstr/ggalt")
library(dplyr)   # for data_frame() & arrange()

df <- data_frame(country=c("Germany", "France", "Vietnam", "Japan", "Poland", "Lebanon",
                           "Australia", "South\nKorea", "Canada", "Spain", "Italy", "Peru",
                           "U.S.", "UK", "Mexico", "Chile", "China", "India"),
                 ages_35=c(0.39, 0.42, 0.49, 0.43, 0.51, 0.57,
                           0.60, 0.45, 0.65, 0.57, 0.57, 0.65,
                           0.63, 0.59, 0.67, 0.75, 0.52, 0.48),
                 ages_18_to_34=c(0.81, 0.83, 0.86, 0.78, 0.86, 0.90,
                                 0.91, 0.75, 0.93, 0.85, 0.83, 0.91,
                                 0.89, 0.84, 0.90, 0.96, 0.73, 0.69),
                 diff=sprintf("+%d", as.integer((ages_18_to_34-ages_35)*100)))

df <- arrange(df, desc(diff))
df$country <- factor(df$country, levels=rev(df$country))

percent_first <- function(x) {
  x <- sprintf("%d%%", round(x*100))
  x[2:length(x)] <- sub("%$", "", x[2:length(x)])
  x
}

gg <- ggplot()
gg <- gg + geom_segment(data=df, aes(y=country, yend=country, x=0, xend=1), color="#b2b2b2", size=0.15)
gg <- gg + geom_segment(data=df, aes(y=country, yend=country, x=ages_35, xend=ages_18_to_34), color="#b2b2b2", size=2)
gg <- gg + geom_point(data=df, aes(x=ages_35, y = country), size=3, colour = "#9fb059")
gg <- gg + geom_point(data=df, aes(x=ages_18_to_34, y = country), size=3, colour = "#edae52")
gg <- gg + geom_text(data=dplyr::filter(df, country=="Germany"),
                     aes(x=ages_35, y=country, label="Ages 35+"),
                     color="#9fb059", size=3, vjust=-1, fontface="bold", family="Calibri")
gg <- gg + geom_text(data=dplyr::filter(df, country=="Germany"),
                     aes(x=ages_18_to_34, y=country, label="Ages 18-34"),
                     color="#edae52", size=3, vjust=-1, fontface="bold", family="Calibri")
gg <- gg + geom_text(data=df, aes(x=ages_35, y=country, label=percent_first(ages_35)),
                     color="#9fb059", size=2.75, vjust=1.75, family="Calibri")
gg <- gg + geom_text(data=df, color="#edae52", size=2.75, vjust=1.75, family="Calibri",
                     aes(x=ages_18_to_34, y=country, label=percent_first(ages_18_to_34)))
gg <- gg + geom_rect(data=df, aes(xmin=1.05, xmax=1.175, ymin=-Inf, ymax=Inf), fill="#efefe3")
gg <- gg + geom_text(data=df, aes(label=diff, y=country, x=1.1125), fontface="bold", size=3, family="Calibri")
gg <- gg + geom_text(data=filter(df, country=="Germany"), aes(x=1.1125, y=country, label="DIFF"),
                     color="#7a7d7e", size=3.1, vjust=-2, fontface="bold", family="Calibri")
gg <- gg + scale_x_continuous(expand=c(0,0), limits=c(0, 1.175))
gg <- gg + scale_y_discrete(expand=c(0.075,0))
gg <- gg + labs(x=NULL, y=NULL, title="The social media age gap",
                subtitle="Adult internet users or reported smartphone owners who\nuse social networking sites",
                caption="Source: Pew Research Center, Spring 2015 Global Attitudes Survey. Q74")
gg <- gg + theme_bw(base_family="Calibri")
gg <- gg + theme(panel.grid.major=element_blank())
gg <- gg + theme(panel.grid.minor=element_blank())
gg <- gg + theme(panel.border=element_blank())
gg <- gg + theme(axis.ticks=element_blank())
gg <- gg + theme(axis.text.x=element_blank())
gg <- gg + theme(plot.title=element_text(face="bold"))
gg <- gg + theme(plot.subtitle=element_text(face="italic", size=9, margin=margin(b=12)))
gg <- gg + theme(plot.caption=element_text(size=7, margin=margin(t=12), color="#7a7d7e"))
gg
```

# Analysis: descriptive statistics
```{r }
hsb <- read_dta('c:/chuck/NYU/Stat Group/R Workshop/hsb.dta')
summary(hsb)
```

# Analysis: frequencies
```{r }
library(descr)
with(bank, freq(JOBCAT, plot=FALSE))
```

# Analysis: crosstabulation
```{r echo=FALSE, results='hide'}
library(descr)
ctab <- with(bank, crosstab(SEX, JOBCAT %in% c("Clerical","Office trainee"), 
                    prop.r=TRUE, plot=FALSE, chisq=TRUE))
```

```{r echo=TRUE, results='hide'}
library(descr)
with(bank, CrossTable(SEX, JOBCAT %in% c("Clerical","Office trainee"), 
                    prop.r=TRUE, prop.chi=FALSE, prop.t=FALSE, prop.c=FALSE, chisq=TRUE))
```

```{r , echo=FALSE, comment="."}
ctab
```

# More detailed descriptive statistics for quantitative variables
```{r }
library(psych)
psych::describe(subset(hsb, select=c(read,write,math,science,socst)))
```

# Analysis: t-tests
- Following these examples: http://www.ats.ucla.edu/stat/stata/output/ttest_output.htm
- One sample t-test where hypothesized mean for write score is 50
```{r }
hsb <- read_dta('c:/chuck/NYU/Stat Group/R Workshop/hsb.dta')
with(hsb, t.test(write, mu=50))
```

- Paired t-test comparing read and write scores
```{r }
with(hsb, t.test(write, read, paired = TRUE))
```

- Independent-samples t-test comparing write scores of male and female students
```{r }
t.test(write ~ female, data = hsb, var.equal = TRUE)
```

- Do not assume equal variance in each group
```{r }
t.test(write ~ female, data = hsb)
```

# Analysis: Correlation
```{r }
hsb <- read_sav("c:/chuck/NYU/Stat Group/R Workshop/hsb2.sav")
round(cor(subset(hsb, select=c(READ,WRITE,MATH,SCIENCE,SOCST))), 2)
cor.test(~ READ + SOCST, data = hsb)
```

# Analysis: Linear Regression
- Following this SPSS example: http://www.ats.ucla.edu/stat/spss/output/reg_spss.htm
```{r }
hsb <- read_sav("c:/chuck/NYU/Stat Group/R Workshop/hsb2.sav")
head(hsb)
```
- Do the multiple regression analysis using the R lm() function (for linear model)
```{r }
lm.1 <- lm(SCIENCE ~ MATH + FEMALE + SOCST + READ, data = hsb)
```

- Summarize the fitted model
```{r }
summary(lm.1)
```

- Confidence intervals
```{r }
confint(lm.1)
```

# Analysis: Logistic Regression
- Following this SPSS example: https://stats.idre.ucla.edu/spss/examples/alr2/applied-logistic-regression-second-edition-by-hosmer-and-lemeshowchapter-2-multiple-logistic-regression/
- Chapter 2, Applied Logistic Regression, Second Edition, by Hosmer and Lemeshow

Variable | Description                               | Codes/Values
---------|-------------------------------------------|------------------------------
LOW      | Low Birth Weight                          | 1 = BWT<=2500g, 0 = BWT>2500g
AGE      | Age of Mother                             | Years
LWT      | Weight of Mother at Last Menstrual Period | Pounds
RACE     | Race                                      | 1=White, 2=Black, 3=Other
FTV      | First Trimester Physician Visits          | 0=None, 1=One, 2=Two,etc.

- R has a function, glm(), for generalized linear models, and logistic regression is a special case of generalized linear models
- This is similar to the functionality of PROC GENMOD in SAS
- Before modeling, we will read in the data and convert some variables to factors

```{r }
lowbwt <- read_sav('c:/chuck/NYU/Stat Group/R Workshop/lowbwt.sav')
lowbwt$RACE <- car::recode(lowbwt$RACE, "1='White';2='Black';3='Other'", as.factor = TRUE)
lowbwt$RACE <- relevel(lowbwt$RACE, ref = 'White')
```

- And now do the logistic regression

```{r }
lrm.1 <- glm(LOW ~ AGE + LWT + RACE + FTV, family='binomial', data = lowbwt)
```

- Summarize it
```{r }
summary(lrm.1)
```

- Odds ratios and confidence intervals
```{r }
exp(coef(lrm.1)[-1]) # Removing constant
exp(confint(lrm.1)[-1,]) # Removing constant from confidence intervals too
```

# Extracting Data from Google Scholar
```{r}
library(scholar)
library(purrr)

DF <- data.frame(Person = c("Chuck Cleland","Sam Friedman","Holly Hagan","Dani Ompad"),
                 GID = c("AuiR_rAAAAAJ","bQwCIGQAAAAJ","67LUQJUAAAAJ","Zxdn7DwAAAAJ"))

do.call(rbind, map(map(DF$GID, get_profile), 
                   function(x){data.frame(Person = x$name,
                                          Total.Cites = x$total_cites,
                                          h.index = x$h_index,
                                          i10.index = x$i10_index)}))
```

# Generating Non-Random Data
```{r}
seq(1, 10, 2)
104:114
rep(c('A','B','C'), 3)
rep(c('A','B','C'), each = 3)
seq(1, 4, len = 6)
expand.grid(Person = 1:5, Visit = 1:3)
```

# Generating Random Data
```{r}
rnorm(10)
runif(10, min=100,max=999)
rbinom(10, 1, .5)
sample(letters[1:3], size = 30, replace=TRUE, prob = c(.8,.1,.1))
sample(10)
```

# More on R Packages
- There is an App for that
- In R, there is a package for that
- ggplot2
     + plots
- dplyr
     + filtering
     + subsetting
     + variable creation
     + summarizing
- tidyr
     + reshape wide to long
     + reshape long to wide
- openxlsx
     + read and write MS-Excel files
- descr
     + frequency tables and crosstabs
- lme4
     + linear and generalized linear mixed effects

# Where and how to find R packages
- Comprehensive R Archive Network (CRAN; https://cran.r-project.org/mirrors.html)
- CRAN task views (https://cloud.r-project.org/web/views/)
- Google (e.g., "r-project psychometrics")
- http://www.rdocumentation.org/
- R <code>available.packages()</code> function

# How to install and load R packages
- Install
     + R <code>install.packages()</code> function
     + Part of the RStudio and RWindows menus
- Load a package to make content available to R
     + R <code>library()</code> function
```{r eval=FALSE}
install.packages("epiR")
```
```{r }
library(epiR)
head(ls("package:epiR"), n = 30)
```

# Resources to Learn More
- http://www.r-project.org/
- http://www.rdocumentation.org/
- Springer Use R! Series (http://www.springer.com/series/6991)
- Cross Validated (http://stats.stackexchange.com/)

